Predicting Remaining Useful Life of NASA Aircraft Engines

### 2025-03-21 index.qmd
updates

Cox proportional hazards (CPH) models are statistical regression models
specializing in modeling time-to-event predictions with survival data.
Survival data is data with a value for time and an event
[@Abeysekera2009]. For example, in a study investigating the survival
likelihood of mechanical equipment, the time value would begin when the
equipment begins functioning and would conclude in the event the
equipment stops functioning [@Smith2003]. CPH models can use single or
multiple covariates as a means of creating predictions until failure and
are capable of dealing with survival data that contains censored data as
well [@Seung2023]. Censored data is when the information about an
individual in a study is only known for a certain period of time, such
as the length of the study [@klein2005, 1]. The models also have the
advantage of being semi-parametric which allows more flexibility as
compared to other models [@Ming-Chiang2014].

@@

An additional useful equations is the cumulative hazard function. The
cumulative hazard function can be defined as

@@

In order to use CPH to model time to event survival data, the data must
contain a value for time, a binary field indicating the event being
modeled, an id field, and one or multiple covariate fields. The time
value is typically formatted as a time series for each subject in the
dataset. Subjects will have a time value from the beginning of the
observation period, to an arbitrary time that may or may not be the time
the event being modeled for occurs. Data is considered censored when the
subject is not given the opportunity to experience the event in the
observation period and is handled differently by the model. The
covariate fields in the dataset can be categorical or numerical
[@nahhas2025].

After verifying the survival data is formatted properly, the CPH model
can be created in R. Important steps to consider are choosing covariates
and verifying the model meet the model's assumptions after the model is
created.

#### Choosing covariates

From the study the data is from, we know that some of the sensors will
not play a role in the RUL. As such, we will run a univariate test of
all covariates to determine which ones will actually impact the RUL. We
will use the p-value provided by the Wald test with a significance level
of $\alpha = 0.05$.

#### Generating Model

@@

When creating a CPH model, to ensure an accurate model is created the
following assumptions must be met. If they are not met, the data must
either be transformed, stratified, or other covariates may need to be
selected for the model. The assumptions associated with the CPH model
are defined below.

#### Proportional Hazards Assumption

The proportional hazards assumption has been proven to be critical to
evaluate prior to using CPH when modeling survival data. The
proportional hazards assumption states that the ratio of hazard rates
for any two subjects must be constant over time regardless of the point
in time of the measurement. This means the effect of the covariates must
be constant over time. [@Bustan2018]

This assumption can be tested using the R package `survival`. The
package contains a proportional hazards test called `cox.zph()`
[@R-base]. Using the hypothesis,

@@

```{r, echo=FALSE}
example_surv_object <- Surv(time = lung$time, event = lung$status)

example_coxmodel <- coxph(example_surv_object ~ meal.cal + age, data = lung)

example_phtest <- cox.zph(example_coxmodel)

@@

#### Linearity Assumption

The linearity assumption states that the relationship between covariates
and the log of the hazard rate must be linear. This assumption can be
checked by plotting Martingale residuals against the range of values
present in each covariate.

Martingale residuals are a type of residuals used in survival analysis
that show the discrepancy between the observed and expected number of
events. $MartingaleResiduals = ObservedEvents - ExpectedEvents$

Residuals can first be found by using the `resid(type = "martingale")`
function and passing martingale in the type argument creating a vector
of residuals that can be plotted against the range of values for each
covariate.

If the resulting residuals are linear and appear to have a slope of
zero, this assumption is not violated.[@Amini2015]

Below is an example plot showing Martingale residuals plotted against
the range of values in the covariate being tested. From the plot, it is
clear there is no clear slope in the residuals indicating the linearity
assumption is not broken. For a model with multiple covariates, multiple
plots will need to be made.

#### Independent Censoring Assumption

This assumption describes how if an event is censored it should not be
considered when the model is being created. Censored data is difficult
to accommodate when Cox proportional hazards models are used in other
industries, such as the medical field, because observation times for the
subjects involved in studies may have decades before an event is
observed and the event may not have the opportunity to occur during the
study period. In this scenario, it would not be appropriate to consider
this subject when modeling survival.

#### Independence Assumption

Survival times of observed subjects should be independent of each other.
In an example where survival of plane engines were being evaluated, if
each plane has two engines and the failure of one engine caused the
accelerated failure of another due to increased workload, it would
produce biased conclusions and violate the independence assumption.

>>>>

Cox proportional hazards (CPH) models are statistical regression models
specializing in modeling time-to-event predictions with survival data.
Survival data is data with a value for time and an event
[@Abeysekera2009]. For example, in a study investigating the survival
likelihood of mechanical equipment, the time value would begin when the
equipment begins functioning and would conclude in the event the
equipment stops functioning [@Smith2003]. CPH models can use single or
multiple covariates as a means of creating predictions until failure and
are capable of dealing with survival data that contains censored data as
well [@Seung2023]. Censored data is when the information about an
individual in a study is only known for a certain period of time, such
as the length of the study [@klein2005]. The models also have the
advantage of being semi-parametric which allows more flexibility as
compared to other models [@Ming-Chiang2014].

@@

An additional useful equation is the cumulative hazard function. The
cumulative hazard function can be defined as

@@

In order to use CPH to model time to event survival data, the data must
contain a value for time, a binary field indicating the event being
modeled, an id field, and one or multiple covariate fields that can be
categorical or numerical. The time value is typically formatted as a
time series for each subject in the dataset. The first entry for
subjects will be a time value from the beginning of the observation
period with the final entry time varying by subject based on if the
event occurred for that individual or if the study ended before the
event occurred. Subjects who did not have the event occur by the end of
the observation period are considered censored entries and will be
handleded different in the model [@nahhas2025].

### Covariate Selection

Depending on the study, there will be covariates that do not impact the
probability of survival. Statistical methods such as the Wald Test can
be used to determine which ones to use in the final CPH model.
Generating a univariate CPH model for each covariate and using the Wald
test with a significance level of $\alpha = 0.05$ will allow for
determination of if a covariate is significant. Any covariates that are
determined to be not significant to the survival probability can be
removed from the final CPH model.

@@

```{r, echo = FALSE}
#generation of the model for sample plots
example_surv_object <- Surv(time = lung$time, event = lung$status)

example_coxmodel <- coxph(example_surv_object ~ meal.cal + age, data = lung)
```

There are four assumptions that a CPH model must meet in order to be an
accurate model. For any violation of those assumptions, the data will
need to be either transformed, stratified, or other covariates may need
to be selected for the model. The four assumptions are the independence
assumption, non-informative censoring, linearity assumption, and
proportional hazards assumption.

**Independence Assumption**

CPH assumes that the survival times of observed subjects are independent
of each other. This is similar to most regression models, though CPH
does not assume that the residuals of the model be normally distributed
or have constant variance [@nahhas2025].

**Non-informative Censoring Assumption**

CPH assumes that censoring is non-informative which means that each
subject has the same risk of experiencing the event regardless of one of
them being censored. Thus, the knowledge of censoring does not result in
any new information being provided [@nahhas2025].

**Linearity Assumption**

CPH assumes the relationship between covariates and the outcome is a
linear relationship [@nahhas2025]. For CPH, the outcome for that
relationship is the log of the hazard rate. This assumption is checked
visually using the Martingale residuals for each covariate
[@nahhas2025]. Martingale residuals are a type of residauls used in
survival analysis that show the discrepancy between the observed and the
expected number of events. The equation is
$\text{Martingale Residuals} = \text{Observed Events} - \text{Expected Events}$.
The resulting residuals are examined visually and if they are linear and
appear to have a slope of zero, the linearity assumption is not violated
[@Amini2015].

Below is an example plot showing Martingale residuals plotted for a
covariate being tested. From the plot, it is visually clear that the
residuals are linear and appear to have a slope of zero. For this
covariate then, the linearity assumption has not been violated. If a
model is being generated with multiple covariates, multiple Martingale
residual plots will need to be generated in order to verify this
assumption for each covariate.

**Proportional Hazards Assumption**

The major assumption for CPH is the proportional hazards assumption
which states that the ratio of hazard rates for any two subjects must be
constant at all times. This means the effect of the covariates must be
constant over time [@Bustan2018]. To test this assumption, the
hypothesis test below is used along with Schoenfeld partial residuals.

@@

```{example_phtest <- cox.zph(example_coxmodel)}

@@
